### 2020-09-24 latex/vlp.tex
Finished the first edit of VLP

\section{The VLP Constraint}
% Go through the development of the constraint and prove it gains energy
To motivate why a VNHC could inject energy into the VLP in a human-like
manner, we will examine the motion of a person standing on a swing.
As can be seen in Figure \ref{fig:child-vlp}, their center of mass moves closer
to the swing's pivot when they stand, and moves away from the pivot when they
squat.
This is equivalent to the VLP model from Figure \ref{fig:vlp-model}.
\begin{figure}
   \centering
   \begin{subfigure}[t]{0.45\textwidth}
      \includegraphics[]{images/child_vlp_standing.png}
      \caption{A person standing on a swing has their center of mass 
      close to the pivot.}
   \end{subfigure}
   \hfill
   \begin{subfigure}[t]{0.45\textwidth}
      \includegraphics[]{images/child_vlp_squatting.png}
      \caption{When a person squats on a swing, their center of mass extends
      away from the pivot.}
   \end{subfigure}
   \caption{The VLP representation of a person on a standing swing.}
   \label{fig:child-vlp}
\end{figure}

@@

This optimal pumping strategy is straightforward:
one stands at the lowest point of the swing, and squats at the highest point.
Looking at the VLP representation, the pendulum shortens at the bottom of the
swing, and lengthens at the top. 
For an intuitive explanation, conservation of angular momentum indicates that
shortening the pendulum at the bottom forces the mass to gain speed to
compensate for the reduced length \cite{how_to_pump_a_swing}.
Energy is not conserved in this process, so the pendulum gains kinetic energy
and reaches a higher point at the peak of its swing.
Lengthening the pendulum when it reaches this peak means gravity
imparts a larger angular momentum to the mass by the time it reaches the bottom
of its swing, which in turn is converted to a higher velocity when the
pendulum is shortened.
By alternating these processes, the pendulum experiences a net gain in
rotational energy on average.

The time-optimal controller described by \cite{pumping_swing_standing_squatting}
depends on a trajectory in time. We will convert this 
into a time-independent controller derived from the time-optimal
strategy. Notice that the optimal controller depends on knowing when the
system is at the ``bottom" or ``top" of the swing. Since being at the ``bottom"
is the same as saying the pendulum angle is \(q = 0\) and being at the ``top" is
equivalent to the pendulum's momentum being \(p = 0\), the time-independent
controller will necessarily use the full phase \((q,p)\). While VHCs cannot
satisfy this requirement, our method of VNHCs is especially well suited to this
problem.

Figure \ref{fig:vlp-optimal-controller-qp-plane} shows the time-optimal strategy directly
on the \((q,p)\)-plane.
Since the VLP must stay the same length inside each quadrant and
changes length only when it crosses one of the axes, we
will use the angle \(\theta = \arctan_2(p,q)\) in the \((q,p)\)-plane to define
the optimal controller \(l^\star(\theta)\) (\ref{eqn:vlp-optimal-controller}).
Figure \ref{fig:vlp-optimal-controller} shows this time-optimal controller as a function of
\(\theta\).
\begin{equation}\label{eqn:vlp-optimal-controller}
   l^\star(\theta):= \begin{cases}
      \underline{l} & \theta \in [-pi, -\frac{\pi}{2}[ \cup [0,\frac{\pi}{2}[ \\
      \overline{l} & \theta \in [-\frac{\pi}{2},0[ \cup [\frac{\pi}{2}, \pi[ \\
   \end{cases}
\end{equation}

\begin{figure}
   \centering
   \begin{subfigure}[t]{0.45\textwidth}
      \includestandalone[width=\linewidth]{images/vlp_optimal_controller_qp_plane}
      \caption{The time-optimal controller mapped to the \((q,p)\)-plane.}
      \label{fig:vlp-optimal-controller-qp-plane}
   \end{subfigure}
   \hfill
   \begin{subfigure}[t]{0.45\textwidth}
      \includestandalone[width=\linewidth]{images/vlp_optimal_controller}
      \caption{The time-optimal controller converted to a VNHC-style 
      function \(l^\star(\theta)\), where \(\theta := \arctan_2(p,q)\).}\label{fig:vlp-optimal-controller}
   \end{subfigure}
   \caption{The time-optimal controller for a standing swing as derived by
      \cite{pumping_swing_standing_squatting}. The colour red
      corresponds to standing and blue to squatting.}
\end{figure}

To find a continuously differentiable function which approximates this controller,
we can attach sinusoids at the transition points of
\(l^\star(\theta)\) (see Figure \ref{fig:vlp-T-controller}). Let
\(\Delta l := (\overline{l} - \underline{l})/2\) and 
\(l_{\text{avg}} := (\overline{l} + \underline{l})/2\).
Providing these newly attached sinusoids with a frequency \(\omega = \frac{pi}{T}\) 
(where \(T \in ]0,\frac{\pi}{2}]\) is a control parameter), we have a new
controller \(l_T(\theta)\):
\begin{equation}\label{eqn:vlp-T-controller}
   l_T(\theta) = \begin{cases}
      \underline{l} & \theta \in \left[-\pi + \frac{T}{2}, -\frac{\pi}{2} - \frac{T}{2}\right] 
      \cup \left[\frac{T}{2}, \frac{\pi}{2} - \frac{T}{2}\right] \\
      \overline{l} & \theta \in \left[-\frac{\pi}{2} + \frac{T}{2}, -\frac{T}{2}\right] 
      \cup \left[\frac{\pi}{2} + \frac{T}{2}, \pi - \frac{T}{2}\right] \\
      -\Delta l \sin(\omega(\theta + \pi)) + l_{\text{avg}} & \theta \in
      \left[-\pi,-\pi + \frac{T}{2}\right] \\
      -\Delta l \sin(\omega \theta) + l_\text{avg} & \theta \in [-\frac{T}{2},
      \frac{T}{2}] \\
      \Delta l \sin(\omega(\theta - a)) + l_\text{avg} & 
      \theta \in \left[a - \frac{T}{2}, a + \frac{T}{2}\right] \text{ for } 
      a \in \left\{-\frac{\pi}{2}, \frac{\pi}{2}\right\} \\
      -\Delta l \sin(\omega(\theta-\pi)) & \theta \in \left[\pi - \frac{T}{2},\pi\right] \\
   \end{cases}
\end{equation}

@@

This controller approximates the optimal controller, since 
\[
   \lim\limits_{T \rightarrow 0} l_T(\theta) = l^\star(\theta)
\]
Unfortunately, while \(l_T(\theta)\) is continuously differentiable, it is not
twice continuously differentiable for most values of \(T\).
If we wish to use it as a VNHC, we must ensure that either the generalized
forces \(\tau\) acting on \(p_l\) can be discontinuous, or we must find a value
of \(T\) where this controller is smooth.
Thankfully, \(l_{\frac{\pi}{2}}(\theta)\) is smooth; it can be simplified into
the expression (\ref{eqn:vlp-smoothed-controller}), and is plotted for
demonstration in Figure \ref{fig:vlp-smoothed-controller}.
\begin{equation}\label{eqn:vlp-smoothed-controller}
   l_\frac{\pi}{2}(\theta) = -\Delta l \sin(2\theta) + l_{\text{avg}}
\end{equation}

\begin{figure}
   \centering
   \includestandalone[width=0.8\textwidth]{images/vlp_smoothed_controller}
   \caption{The smoothed VLP controller which corresponds to the VNHC \(l =
      l_\frac{\pi}{2}(\theta)\).}\label{fig:vlp-smoothed-controller}
\end{figure}

We therefore set our VNHC to be
\[
   h(q,p) = l - l_\frac{\pi}{2}\left(\theta(q,p)\right)
\]

This VNHC is not the same as the optimal controller
(\ref{eqn:vlp-optimal-controller}), so we need to prove it will still gain
energy. We will do so by showing the dynamics (\ref{eqn:vlp-hamiltonian}) of the
VLP diverge from the origin, regardless of the initial condition of the system. 
To do this, we will use the notion of an anti-Lyapunov function.

\begin{defn}\label{defn:anti-lyapunov}
   Let \(\dot{x} = f(x) \in R^n\) be an ODE with \(f(0) = 0\). 
   A positive definite function \(V(x)\) is \textbf{anti-Lyapunov} if 
   \[
      \dot{V}(x) = dV(x) f(x) \geq 0
   \]
   where the inequality is strict for almost every \(x \in \R^n\).
\end{defn}

By applying Lyapunov stability theory \cite{lyapunov},
we see that an anti-Lyapunov function
\(V\) is simply a Lyapunov function for the negative-time system 
\(\dot{x} = -f(x)\). In other words, since \(V\) proves that the orbit \(x(-t)\) 
of the system travelling backwards in time is asymptotically converging to the
equilibrium, the system must be diverging from the equilibrium in forward time.
If we can find such a function for the VLP's dynamics under our VNHC, the path
generated by the VNHC in the \((q,p)\)-plane must be diverging from the origin
so long as the initial condition is not an equilibrium.
This in turn tells us the magnitude of the momentum \(p\) is increasing every
time the path hits the \(p\)-axis, which means the VLP is, on average,
increasing in energy.

We will show that anti-Lyapunov function exists for our VNHC.
In order to do so, we require the following lemma.

\begin{lemma}\label{lemma:sign-of-cube}
   For any \(x,y \in \R\):
   \[
      \sign{x^3 - y^3} = \sign{x-y}
   \]
\end{lemma}
\begin{proof}
   Observe first that
   \[
      x^3 - y^3 =  (x-y)(x^2 + xy + y^2)
   \]
   Since \( (x-y)^2 = x^2 - 2xy + y^2) \geq 0 \), 
   \[
      \frac{x^2 + y^2}{2} \geq xy
   \]
   which means \(x^2 + xy + y^2 \geq 0\), proving the lemma.
\end{proof}

\begin{thm}\label{thm:vlp-energy-stabilization}
   Take the VLP with Hamiltonian dyanmics (\ref{eqn:vlp-hamiltonian}) and define
   \(\theta := \arctan_2(p,q)\). Suppose the initial condition is not one of the
   equilibria: \((q(0),p(0)) \not \in \left\{(0,0), (\pm\pi,0)\right\}\).
   A VNHC of the form \(h(q,p) = l - l(\theta)\) injects energy if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l(\theta) - l_\text{avg}\right)sin(2\theta) \leq 0 \text{ }\forall \theta \in \mathbb{S}^1
   \end{equation}
   with the property that the inequality is strict for almost every \(\theta\).
   Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) leads to
   energy dissipation.
\end{thm}
\begin{proof}
    Choose, as a candidate anti-Lyapunov function, the energy for the
    average-length pendulum (with zero potential at the bottom of the swing)
    \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
                    + m g l_\text{avg} (1-\cos(q))
    \]
    which is positive definite at \((0,0)\) and has derivative 
    \[
      \dot{E}_\text{avg} = \frac{-g\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)}
                 {l_\text{avg}^2l(\theta)^2}
    \]
    Observe that \(\sign{\sin(q)p} = \sign{\sin(2\theta)}\) and, 
    by Lemma \ref{lemma:sign-of-cube},
    \[ 
       \sign{l(\theta)^3 - l_\text{avg}^3} = \sign{ l(\theta) - l_\text{avg}}
    \]
    Then the derivative of \(E_\text{avg}\) is almost always positive, since
    \begin{align*}
       \label{eq:}
       \sign{\dot{E}_\text{avg}} &= \sign{-\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)} \\
                   &= -\sign{\sin(2\theta) \left(l(\theta) - l_\text{avg}\right)} \\
                   &\geq 0 \text{ (by assumption)}
    \end{align*}
    Hence, \(E_\text{avg}\) is an anti-Lyapunov function.
    Since the initial condition is not an equilibrium,
    the orbit of the VNHC in the \((q,p)\) plane is diverging away from the
    origin and the VLP is gaining energy on average.
    Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
    means \(E_\text{avg}\) is a regular Lyapunov function, since its derivative
    will be negative almost everywhere.
\end{proof}

\begin{cor}
   The VNHC \(l = l_\frac{\pi}{2}(\theta)\) injects energy into the VLP
   over time, since 
   \[
      \left(l_\frac{\pi}{2}(\theta) - l_\text{avg}\right)\sin(2\theta) = 
   -\Delta l \sin^2(2\theta) \geq 0
   \] 
   Likewise, let 
   \(l\inv_\frac{\pi}{2}(\theta) := l_\text{avg} - \Delta l \sin(2\theta)\).
    The VNHC \(h(q,p) = l - l\inv_\frac{\pi}{2}(\theta)\) which flips 
    \(l_\frac{\pi}{2}(\theta)\) about \(l_\text{avg}\) will dissipate energy from
   the VLP, since 
   \[
      \left(l\inv_\frac{\pi}{2}(\theta) - l_\text{avg}\right)\sin(2\theta) = 
      -\Delta l \sin^2(2\theta) \leq 0
   \] 
\end{cor}

The types of VNHCs which satisfy Theorem \ref{thm:vlp-energy-stabilization} are illustrated by Figure
\ref{fig:vlp-energy-in-out}. To stabilize specific energy level sets, one simple
approach is to switch between injection and dissipation VNHCs depending on the
current energy level.
For the VNHCs we designed in this section, this means toggling
between \(l_\frac{\pi}{2}(\theta)\) and \(l\inv_\frac{\pi}{2}(\theta)\),
possibly with some hysteresis to avoid infinite switching.

@@

Let us define \((l^\star)\inv(\theta)\) by swapping the order of 
\(\underline{l}\) and \(\overline{l}\) in \(l^\star(\theta)\). Since this 
\textit{minimizes} the derivative of \(E_\text{avg}\) under the restriction
\(l(\theta) \in [\underline{l},\overline{l}]\), we can predict that 
\((l^\star)\inv(\theta)\) is the VNHC representation of an optimal 
energy dissipation controller.
This is, in fact, true: \cite{pumping_swing_standing_squatting} showed 
that squatting at the lowest point of a swing and standing at the highest
point (instead of standing and squatting resp.) produces the
time-optimal trajectory for stopping a standing swing.

All together, these results show that VNHCs can replicate the time-optimal
strategy performed by humans in a time-independent manner. Furthermore, we see
that VNHCs are a powerful tool for creating simple energy stabilization
techniques based on natural human motion.

>>>>

\section{The VLP Constraint}
% Go through the development of the constraint and prove it gains energy
To motivate why a VNHC could inject energy into the VLP in a human-like
manner, we will examine a person standing on a swing.
As can be seen in Figure \ref{fig:child-vlp}, their center of mass moves closer
to the swing's pivot when they stand, and moves away from the pivot when they
squat.
This is equivalent to the VLP model from Figure \ref{fig:vlp-model}, where
standing and squatting correspond to shortening and lengthening the pendulum
(respectively).
\begin{figure}
   \centering
   \begin{subfigure}[t]{0.45\textwidth}
      \includegraphics[]{images/child_vlp_standing.png}
      \caption{A person standing on a swing has their center of mass 
      close to the pivot.}
   \end{subfigure}
   \hfill
   \begin{subfigure}[t]{0.45\textwidth}
      \includegraphics[]{images/child_vlp_squatting.png}
      \caption{When a person squats on a swing, their center of mass extends
      away from the pivot.}
   \end{subfigure}
   \caption{The VLP representation of a person on a standing swing.}
   \label{fig:child-vlp}
\end{figure}

@@

A child's optimal pumping strategy is straightforward:
they stand at the lowest point of the swing, and squat at the highest point.
Looking at the VLP representation, the pendulum shortens at the bottom of the
swing, and lengthens at the top. 
For an intuitive explanation, conservation of angular momentum indicates that
shortening the pendulum at the bottom forces the mass to gain speed to
compensate for the reduced length \cite{how_to_pump_a_swing}.
Energy is not conserved in this process, so the pendulum gains kinetic energy
and reaches a higher point at the peak of its swing.
Lengthening the pendulum when it reaches this peak means gravity
imparts a larger angular momentum to the mass by the time it reaches the bottom
of its swing, which in turn is converted to a higher velocity when the
pendulum is shortened.
By alternating these processes, the pendulum experiences an average net gain in
rotational energy.

Notice that the child's pumping strategy requires knowledge of
when the system is at the ``bottom" or ``top" of the swing. 
Since being at the ``bottom" is equivalent to having angle \(q = 0\) and being
at the ``top" is equivalent to having momentum \(p = 0\), a controller based on
this strategy will necessarily use the full phase \((q,p)\). 
This is why we will use VNHCs instead of VHCs.

The time-optimal controller from \cite{pumping_swing_standing_squatting} is (in
our notation)
\[
   l_\text{opt}(q,p) := -\sign{qp}
\]
which is a piecewise-continuous controller that varies between \(\pm 1\). 
This looks similar to a VNHC, but it is not smooth and makes the assumption that 
\(l \in [-1,1]\). 
Since we want our controller emulate realistic human motion, we need to derive a
smooth controller where we allow \(l \in [\underbar{l},\overbar{l}]\); then, we
will prove that it injects energy.

Figure \ref{fig:vlp-optimal-controller-qp-plane} displays \(l_\text{opt}\) 
directly on the \((q,p)\)-plane.
Because the length remains constant inside each quadrant and changes only when it
crosses one of the axes, one can use the angle \(\theta = \arctan_2(p,q)\) to
redefine the time-optimal controller in a way which is easier to ``smooth out".
Figure \ref{fig:vlp-optimal-controller} shows the time-optimal pumping strategy
as a function \(l^\star(\theta)\) (\ref{eqn:vlp-optimal-controller}), where now
the length varies between \([\underbar{l},\overbar{l}]\) rather than 
\([-1,1]\).

\begin{equation}\label{eqn:vlp-optimal-controller}
   l^\star(\theta):= \begin{cases}
      \overline{l} & \theta \in [-\frac{\pi}{2},0[ \cup [\frac{\pi}{2}, \pi[ \\
      \underline{l} & \theta \in [-pi, -\frac{\pi}{2}[ \cup [0,\frac{\pi}{2}[ \\
   \end{cases}
\end{equation}

\begin{figure}
   \centering
   \begin{subfigure}[t]{0.45\textwidth}
      \includestandalone[width=\linewidth]{images/vlp_optimal_controller_qp_plane}
      \caption{The time-optimal controller \(l_\text{opt}\) mapped onto the \((q,p)\)-plane.}
      \label{fig:vlp-optimal-controller-qp-plane}
   \end{subfigure}
   \hfill
   \begin{subfigure}[t]{0.45\textwidth}
      \includestandalone[width=\linewidth]{images/vlp_optimal_controller}
      \caption{The time-optimal controller converted to the alternate
         representation \(l^\star(\theta)\).}
      \label{fig:vlp-optimal-controller}
   \end{subfigure}
   \caption{The time-optimal controller for a standing swing as derived by
      \cite{pumping_swing_standing_squatting}. The colour red
      corresponds to standing, blue to squatting, and 
      \(\theta := \arctan_2(p,q)\) is the angle of the VLP phase in the
      \((q,p)\)-plane.}
\end{figure}

We now define a continuous function which
approximates \(l^\star(\theta)\) by attaching sinusoids at the
discontinuities (see Figure \ref{fig:vlp-T-controller}). 
Let \(\Delta l := (\overline{l} - \underline{l})/2\) and 
\(l_{\text{avg}} := (\overline{l} + \underline{l})/2\).
Providing these newly attached sinusoids with a frequency \(\omega = \frac{\pi}{T}\) 
(where \(T \in \, ]0,\frac{\pi}{2}]\) is a control parameter), we get the
twice-differentiable controller \(l_T(\theta)\) (\ref{eqn:vlp-T-controller}).
\begin{equation}\label{eqn:vlp-T-controller}
   l_T(\theta) = \begin{cases}
      \overline{l} & \theta \in \left[-\frac{\pi}{2} + \frac{T}{2}, -\frac{T}{2}\right] 
      \cup \left[\frac{\pi}{2} + \frac{T}{2}, \pi - \frac{T}{2}\right] \\
      \underline{l} & \theta \in \left[-\pi + \frac{T}{2}, -\frac{\pi}{2} - \frac{T}{2}\right] 
      \cup \left[\frac{T}{2}, \frac{\pi}{2} - \frac{T}{2}\right] \\
      -\Delta l \sin(\omega(\theta + \pi)) + l_{\text{avg}} & \theta \in
      \left[-\pi,-\pi + \frac{T}{2}\right] \\
      -\Delta l \sin(\omega \theta) + l_\text{avg} & \theta \in [-\frac{T}{2},
      \frac{T}{2}] \\
      \Delta l \sin(\omega(\theta - a)) + l_\text{avg} & 
      \theta \in \left[a - \frac{T}{2}, a + \frac{T}{2}\right] \text{ for } 
      a \in \left\{-\frac{\pi}{2}, \frac{\pi}{2}\right\} \\
      -\Delta l \sin(\omega(\theta-\pi)) & \theta \in \left[\pi - \frac{T}{2},\pi\right] \\
   \end{cases}
\end{equation}

@@

This controller approximates the optimal controller because 
\[
   \lim\limits_{T \rightarrow 0} l_T(\theta) = l^\star(\theta)
\]
Unfortunately, while \(l_T(\theta)\) is twice-differentiable, its second
derivative is only piecewise continuous for most values of \(T\).
If we wish to use it as a VNHC, we must ensure that either the generalized
forces \(\tau\) acting on \(p_l\) can be discontinuous (which is certainly
achievable by humans), or we must find a value of \(T\) where this controller is
smooth.
Thankfully, \(l_{\frac{\pi}{2}}(\theta)\) is smooth as it can be simplified into
the expression (\ref{eqn:vlp-smoothed-controller}). 
It is plotted for demonstration in Figure \ref{fig:vlp-smoothed-controller}.
\begin{equation}\label{eqn:vlp-smoothed-controller}
   l_\frac{\pi}{2}(\theta) = -\Delta l \sin(2\theta) + l_{\text{avg}}
\end{equation}

\begin{figure}
   \centering
   \includestandalone[width=0.6\textwidth]{images/vlp_smoothed_controller}
   \caption{The smoothed VLP controller which corresponds to the VNHC \(l =
      l_\frac{\pi}{2}(\theta)\).}\label{fig:vlp-smoothed-controller}
\end{figure}

Because \(l_\frac{\pi}{2}(\theta)\) is smooth and it approximates
\(l^\star(\theta)\), we set our VNHC to be
\[
   h(q,p) = l - l_\frac{\pi}{2}\left(\theta(q,p)\right)
\]
We still need to prove this VNHC will inject energy into a VLP. 
We will do so by showing the dynamics (\ref{eqn:vlp-hamiltonian}) 
diverge from the origin via an ``anti-Lyapunov function".

\begin{defn}\label{defn:anti-lyapunov}
   Let \(\dot{x} = f(x) \in R^n\) be an ODE with \(f(0) = 0\). 
   A positive definite function \(V(x)\) is \textbf{anti-Lyapunov} if 
   \[
      \dot{V}(x) = dV(x) f(x) \geq 0
   \]
   where the inequality is strict for almost every \(x \in \R^n\).
\end{defn}
\textbf{Manfredi: Does this notion already exist elsewhere? I couldn't find any
sources on it, but it seems like it should've been done before.}

By applying Lyapunov stability theory \cite{lyapunov},
we see that an anti-Lyapunov function
\(V\) is simply a Lyapunov function for the negative-time system 
\(\dot{x} = -f(x)\). In other words, \(V\) proves that the orbit \(x(-t)\) 
of the system travelling backwards in time is asymptotically converging to the
origin; hence, the system must be diverging from the origin in forward time.

From Theorem \ref{thm:zero-dynamics}, the dynamics on the constraint manifold of
the VNHC are given by \((\dot{q},\dot{p})\) with 
\(l = l_\frac{\pi}{2}(\theta)\).
Supposing the initial condition is not an equilibrium, the existence of an
anti-Lyapunov function would indicate that the curve traced out on the
\((q,p)\)-plane is diverging from the origin.
This implies that the momentum \(p\) is increasing in magnitude whenever the
curve hits the \(p\)-axis, which in turn means the VLP is gaining energy on
average.

Proving the existence of an anti-Lyapunov function for the VLP
requires the following lemma.

\begin{lemma}\label{lemma:sign-of-cube}
   For any \(x,y \in \R\):
   \[
      \sign{x^3 - y^3} = \sign{x-y}
   \]
\end{lemma}
\begin{proof}
   Observe first that
   \[
      x^3 - y^3 =  (x-y)(x^2 + xy + y^2)
   \]
   The inequality \(x^2 + xy + y^2 \geq 0\) holds because
   \[
      x^2 + xy + y^2 = \left(x  + \frac{y}{2}\right)^2 + \frac{3y^2}{4} \geq 0
   \]
   which proves the lemma.
\end{proof}

\begin{thm}\label{thm:vlp-energy-stabilization}
   Take the Hamiltonian VLP (\ref{eqn:vlp-hamiltonian}) with initial condition 
   \((q_0,p_0) \not \in \left\{(0,0),(\pm\pi,0)\right\}\).
   Define \(\theta := \arctan_2(p,q)\). 
   A VNHC of the form \(h(q,p) = l - l(\theta)\) injects energy if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l(\theta) - l_\text{avg}\right)sin(2\theta) \leq 0 \text{ }\forall \theta \in \mathbb{S}^1
   \end{equation}
   with the property that the inequality is strict for almost every \(\theta\).
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
    Choose, as a candidate anti-Lyapunov function, the energy for the
    average-length pendulum with zero-potential at the bottom of the swing:
    \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
                    + m g l_\text{avg} (1-\cos(q))
    \]
    This is positive definite at \((0,0)\) and has derivative 
    \[
      \dot{E}_\text{avg} = \frac{-g\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)}
                 {l_\text{avg}^2l(\theta)^2}
    \]
    The sign of \(\sin(q)p\) is dependent only on the quadrant in the
    \((q,p)\)-plane, so it is equal to the sign of \(\sin(2\theta)\). 
    Furthermore, by Lemma \ref{lemma:sign-of-cube}:
    \[ 
       \sign{l(\theta)^3 - l_\text{avg}^3} = \sign{ l(\theta) - l_\text{avg}}
    \]
    Thus, the derivative of \(E_\text{avg}\) is almost always positive since
    \begin{align*}
       \label{eq:}
       \sign{\dot{E}_\text{avg}} &= \sign{-\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)} \\
                   &= -\sign{\sin(2\theta) \left(l(\theta) - l_\text{avg}\right)} \\
                   &\geq 0 \text{ (by assumption)}
    \end{align*}
    Therefore \(E_\text{avg}\) is an anti-Lyapunov function.
    Since the initial condition is not an equilibrium,
    the orbit of the VNHC in the \((q,p)\) plane is diverging away from the
    origin and the VLP is gaining energy on average.
    Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
    means \(E_\text{avg}\) is a regular Lyapunov function, so the VLP is
    losing energy while converging to \((0,0)\).
 \end{proof}

\begin{cor}
   Recall that
   \[
      l_\frac{\pi}{2}(\theta) = -\Delta l \sin(2\theta) + l_\text{avg}
   \]
   and define 
   \[
      l^{-}_\frac{\pi}{2}(\theta) := \Delta l \sin(2\theta) + l_\text{avg}
   \]
   The VNHC \(l = l_\frac{\pi}{2}(\theta)\) injects energy into the VLP
   over time, since 
   \[
      \left(l_\frac{\pi}{2}(\theta) - l_\text{avg}\right)\sin(2\theta) = 
      -\Delta l \sin^2(2\theta) \leq 0
   \] 
   The VNHC \(l = l^{-}_\frac{\pi}{2}(\theta)\) dissipates energy
   from the VLP, since 
   \[
      \left(l^{-}_\frac{\pi}{2}(\theta) - l_\text{avg}\right)\sin(2\theta) = 
      \Delta l \sin^2(2\theta) \geq 0
   \] 
\end{cor}

The types of VNHCs which satisfy Theorem \ref{thm:vlp-energy-stabilization} are
illustrated by Figure \ref{fig:vlp-energy-in-out}. 
To stabilize specific energy level sets, a basic approach is to switch
between injection and dissipation VNHCs depending on the current energy level.
For the VNHCs we designed in this chapter, this means toggling
between \(l_\frac{\pi}{2}(\theta)\) and \(l^{-}_\frac{\pi}{2}(\theta)\),
with some hysteresis to avoid infinite switching.

@@

Let us define \((l^\star)^{-}(\theta)\) by swapping the order of 
\(\underline{l}\) and \(\overline{l}\) in \(l^\star(\theta)\). Since this 
\textit{minimizes} the derivative of \(E_\text{avg}\) under the restriction
\(l(\theta) \in [\underline{l},\overline{l}]\), we can predict that 
\((l^\star)^{-}(\theta)\) is the VNHC representation of an optimal 
energy dissipation controller.
This is, in fact, true: \cite{pumping_swing_standing_squatting} showed 
that squatting at the lowest point of a swing and standing at the highest
point (instead of standing and squatting resp.) produces the
time-optimal trajectory for \textit{stopping} a standing swing.

All together, these theory in this chapter shows that VNHCs can replicate the
time-optimal pumping strategy performed by humans. 
Furthermore, we see that VNHCs are a powerful tool for creating simple energy
stabilization techniques based on natural human motion.
